Growth model analysis 3spp WB

### 2019-03-14 MS_Analysis_3spp_4rivers_Patterns_from 2spp.Rmd
switched to bam(), many tests

```{r Load data, message=FALSE}
library(arm)
library(relaimpo)
library(linkedModels)
library(arrayhelpers)
library(tidyverse)
library(gridExtra)

@@

#######################

# 1. Do we need a RE on year in s()? YES
mod0a <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr'), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod1)

mod0 <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod1)

AIC(mod0,mod0a) %>% rownames_to_column() %>% arrange(AIC)

# 2. Best structure for year,species,river
mod1 <- gam(observedLength ~ species + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod2 <- gam(observedLength ~ river + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

@@

# 2-way models
mod3 <- gam(observedLength ~ year * species + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod4 <- gam(observedLength ~ year * riverOrdered * riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod4)

mod5 <- gam(observedLength ~ species * riverOrdered * riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod4)

# mod5 is best
AIC(mod0,mod1,mod2,mod3,mod4,mod5) %>% rownames_to_column() %>% arrange(AIC)

# 3-way models
mod6 <- gam(observedLength ~ year * species * riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

# https://www.fromthebottomoftheheap.net/2017/12/14/difference-splines-ii/
# calculates differences from overall spline for each year
# a lot slower
vB$yearOrdered <- factor(vB$year, levels = 2000:2013)
mod6a <- gam(observedLength ~ year * species * riverOrdered + s(ydayCumul, bs = 'cr', by = (yearOrdered)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

# try with river instead of riverOrdered
mod6b <- gam(observedLength ~ year * species * river + s(ydayCumul, bs = 'cr', by = (yearOrdered)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod6b)

# 3-way is best
AIC(mod1,mod2,mod3,mod4,mod5,mod6,mod6a) %>% rownames_to_column() %>% arrange(AIC)

# Any additive 3-way better? NO
mod7 <- gam(observedLength ~ year + species * riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')))

mod8 <- gam(observedLength ~ year * species + riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')))

mod9 <- gam(observedLength ~ year + species + riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')))

>>>>

```{r Load data, message=FALSE}
library(itsadug)
library(arm)
library(relaimpo)
library(linkedModels)
library(arrayhelpers)
library(tidyverse)
library(gridExtra)

@@

#######################
vB$yearFactor <- factor(vB$year)
vB$yearOrdered <- factor(vB$year, levels = 2000:2013)
vB$yearOrderedSpecies <- factor(paste(vB$yearOrdered,vB$species, sep = '_'))

# 1. Do we need a RE on year in s()? YES, yearOrdered is best
# 
mod0_0 <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr'), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod0_1 <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod0_2 <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr', by = (yearFactor)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod0_3 <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr', by = (yearOrdered)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod0_4 <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr', by = (yearOrderedSpecies)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

AIC(mod0_0,mod0_1,mod0_2,mod0_3,mod0_4) %>% rownames_to_column() %>% arrange(AIC)

# 2. structure of smooths
# with ordered factor, need to include an overall smooth, http://www.sfs.uni-tuebingen.de/~jvanrij/Tutorial/GAMM.html 
mod0_3a <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = (yearOrdered)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod0_3b <- gam(observedLength ~ yearOrdered + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = (yearOrdered)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

# mod0_3b is best, but we want to estimate a trend as in mod0_3a
# add + s(ydayCumul, bs = 'cr') to models
AIC(mod0_3,mod0_3a,mod0_3b) %>% rownames_to_column() %>% arrange(AIC)

# 3. gam vs. bam for this big dataset? bam() is MUCH faster!
system.time( mod0_3a_1 <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = (yearOrdered)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML") )
# 316 sec
summary(mod0_3a_1)

system.time( mod0_3a_2 <- bam(observedLength ~ year + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = (yearOrdered)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML") )
# 8 sec
summary(mod0_3a_2)

# 4. bs = 'cr', vs default 'tp'. AIC for 'cr' is ~ 300 lower. Stick with 'cr'
mod0_3a_2a <- bam(observedLength ~ year + s(ydayCumul) + s(ydayCumul, by = (yearOrdered)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

AIC(mod0_3a_2,mod0_3a_2a)

# 5. Include Individual as a RE??
# Taking a long time, maybe run overnight
vB$tagIndexFactor <- factor(vB$tagIndex)
system.time( mod0_3a_3 <- bam(observedLength ~ year + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = (yearOrdered)) + s(ydayCumul, tagIndexFactor, bs="fs", m=1), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML") )
# sec
compareML(mod0_3a_3,mod0_3a_2)

# 2. Best structure for year,species,river
mod0 <- bam(observedLength ~ year + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = yearOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod1 <- bam(observedLength ~ species + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = yearOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod2 <- bam(observedLength ~ river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = yearOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

@@

# 2-way models
mod3 <- bam(observedLength ~ year * species + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = yearOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod4 <- bam(observedLength ~ year * river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = yearOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod5 <- bam(observedLength ~ species * river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = yearOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

# mod4 is best
AIC(mod0,mod1,mod2,mod3,mod4,mod5) %>% rownames_to_column() %>% arrange(AIC)

# 3-way models
mod6 <- bam(observedLength ~ year * species * river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = yearOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod6)

# Any additive 3-way better? NO
mod7 <- bam(observedLength ~ year + species * river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = yearOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')))

mod8 <- bam(observedLength ~ year * species + river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = yearOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')))

mod9 <- bam(observedLength ~ year + species + river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = yearOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')))
